CrypTool Script \\ Mathematics and Cryptography

### 2002-04-15 script-de/einfuerung_zahlentheorie-de.tex
no message

\author{
(c) B. Esslinger, 1998-2002 \\
Frankfurt am Main \\*[8pt]
}

@@

Der {\em Autor} Bernhard Esslinger m"ochte sich an dieser Stelle bedanken bei den Kollegen in der
Firma und an den Universit"aten Frankfurt, Gie"sen und Karlsruhe, insbesondere
bei Dr. Peer Wichmann vom Forschungszentrum Informatik (FZI) Karlsruhe f"ur
die unkomplizierte Unterst"utzung.

@@

\enlargethispage{0.5cm}
Wie auch bei CrypTool w"achst die Qualit"at des Skripts mit Ihren Anregungen
und Verbesserungen. Wir freuen uns "uber Ihre R"uckmeldung.

@@

\vskip +7 pt \noindent
Die aktuelle Version von CrypTool finden Sie unter \newline
  \href{http://www.CrypTool.de}{\tt http://www.CrypTool.de},~~
  \href{http://www.cryptool.com}{\tt http://www.CrypTool.com}~~ oder ~~
  \href{http://www.cryptool.org}{\tt http://www.CrypTool.org}.
\vskip + 7 pt \noindent
Im Readme zu CrypTool sind die Ansprechpartner f"ur dieses kostenlose Tool genannt.
}

>>>>

\author{
(c) B. Esslinger, 1998-2002 \\
Frankfurt am Main \\
besslinger@web.de \\*[8pt]
}

@@

Der {\em Autor} m"ochte sich an dieser Stelle bedanken bei den Kollegen in der
Firma und an den Universit"aten Frankfurt, Gie"sen und Karlsruhe, insbesondere
bei Dr. Peer Wichmann vom Forschungszentrum Informatik (FZI) Karlsruhe f"ur
die unkomplizierte Unterst"utzung.

@@

\enlargethispage{0.5cm}
Wie auch bei CrypTool w"achst die Qualit"at des Skripts mit Ihren Anregungen
und Verbesserungen. Ich freue mich "uber Ihre R"uckmeldung.

@@

\vskip +7 pt \noindent
Die aktuelle Version von CrypTool finden Sie unter \newline
  \href{http://www.CrypTool.de}{\tt http://www.CrypTool.de},~~
  \href{http://www.cryptool.com}{\tt http://www.CrypTool.com}~~ oder ~~
  \href{http://www.cryptool.org}{\tt http://www.CrypTool.org}.
%\vskip + 7 pt \noindent
%Im Readme zu CrypTool sind die Ansprechpartner f"ur dieses kostenlose Tool genannt.
}

### 2002-04-15 script-de/script-de.tex
no message

Hier werden die Daten mittels symmetrischer Verfahren
verschl"usselt: der Schl"ussel ist ein vom Absender zuf"allig
generierter Sitzungsschl"ussel (session key), der nur f"ur diese
Nachricht verwendet wird. Anschlie"send wird dieser
Sitzungsschl"ussel mit Hilfe des asymmetrischen Verfahrens
verschl"usselt und zusammen mit der Nachricht an den Empf"anger
"ubertragen. Der Empf"anger kann den Sitzungsschl"ussel mit Hilfe
seines geheimen Schl"ussels bestimmen und mit diesem dann die
Nachricht entschl"usseln. Auf diese Weise nutzt man das bequeme
Schl"usselmanagement \index{Schl""usselmanagement} asymmetrischer
Verfahren und kann trotzdem gro"se Datenmengen schnell und effektiv mit
symmetrischen Verfahren verschl"usseln.

@@

\begin{enumerate}
   \item \href{http://www.utm.edu/}{\tt http://www.utm.edu/ }
   \item \href{http://prothsearch.net/index.html}{\tt http://prothsearch.net/index.html}
   \item \href{http://www.mersenne.org/prime.htm}{\tt http://www.mersenne.org/prime.htm }
   \item \href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html }
   \item \href{http://www.eff.org/coop-awards/prime-release1.html}{\tt http://www.eff.org/coop-awards/prime-release1.html }
%   \item \href{http://www.octocad.demon.co.uk/mersenne/prime.htm}{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm }
   \item \href{http://www.informatik.tu-darmstadt.de/TI/LiDIA/}{\tt http://www.informatik.tu-darmstadt.de/TI/LiDIA/ }
   \item \href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{\tt http://www.math.Princeton.EDU/\~{}arbooker/nthprime.html }
   \item \href{http://www.utm.edu/research/primes/notes/by_year.html}{\tt http://www.utm.edu/research/primes/notes/by\_year.html }
   \item \href{http://www.cerias.purdue.edu/homes/ssw/cun}{\tt http://www.cerias.purdue.edu/homes/ssw/cun }
   \item \href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}
   \item \href{http://www.mathematik.ch/mathematiker/goedel.html}{http://www.mathematik.ch/mathematiker/goedel.html}
\end{enumerate}

@@

Ein anderer Name f"ur diese Art von Angriffen ist \index{Man-in-the-middle-attack} \glqq man-­in-­the-­middle-­attack\grqq. Hilfe gegen diese Art
von Angriffen verspricht die Public-­Key­-Zertifizierung, die die \index{Authentizit""at} Authentizit"at "offentlicher Schl"ussel
garantieren soll. Die am weitesten verbreitete Zertifizierungsmethode ist der X.509­-Standard.

>>>>

Hier werden die Daten mittels symmetrischer Verfahren
verschl"usselt: der Schl"ussel ist ein vom Absender zuf"allig
generierter Sitzungsschl"ussel (session key)\index{Session Key}, der nur f"ur diese
Nachricht verwendet wird. Anschlie"send wird dieser
Sitzungsschl"ussel mit Hilfe des asymmetrischen Verfahrens
verschl"usselt und zusammen mit der Nachricht an den Empf"anger
"ubertragen. Der Empf"anger kann den Sitzungsschl"ussel mit Hilfe
seines geheimen Schl"ussels bestimmen und mit diesem dann die
Nachricht entschl"usseln. Auf diese Weise nutzt man das bequeme
Schl"usselmanagement \index{Schl""usselmanagement} asymmetrischer
Verfahren und kann trotzdem gro"se Datenmengen schnell und effektiv mit
symmetrischen Verfahren verschl"usseln.

@@

\begin{enumerate}
   \item \href{http://www.utm.edu/}{\tt http://www.utm.edu/ }
   \item \href{http://prothsearch.net/index.html}{\tt http://prothsearch.net/index.html}
   \item \href{http://www.mersenne.org/prime.htm}{\tt http://www.mersenne.org/prime.htm }
   \item \href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html }
   \item \href{http://www.eff.org/coop-awards/prime-release1.html}{\tt http://www.eff.org/coop-awards/prime-release1.html }
%   \item \href{http://www.octocad.demon.co.uk/mersenne/prime.htm}{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm }
   \item \href{http://www.informatik.tu-darmstadt.de/TI/LiDIA/}{\tt http://www.informatik.tu-darmstadt.de/TI/LiDIA/ }
   \item \href{http://www.math.Princeton.EDU/\~{}arbooker/nthprime.html}{\tt http://www.math.Princeton.EDU/\~{}arbooker/nthprime.html }
   \item \href{http://www.utm.edu/research/primes/notes/by_year.html}{\tt http://www.utm.edu/research/primes/notes/by\_year.html }
   \item \href{http://www.cerias.purdue.edu/homes/ssw/cun}{\tt http://www.cerias.purdue.edu/homes/ssw/cun }
   \item \href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}
   \item \href{http://www.mathematik.ch/mathematiker/goedel.html}{http://www.mathematik.ch/mathematiker/goedel.html}
\end{enumerate}

@@

Ein anderer Name f"ur diese Art von Angriffen ist \index{Man-in-the-middle-attack} \glqq man-­in-­the-­middle-­attack\grqq. Hilfe gegen diese Art
von Angriffen verspricht die Public Key­-Zertifizierung, die die \index{Authentizit""at} Authentizit"at "offentlicher Schl"ussel
garantieren soll. Die am weitesten verbreitete Zertifizierungsmethode ist der X.509­-Standard.

### 2002-04-15 script-gb/einfuerung_zahlentheorie-gb.tex
no message

\title{ CrypTool Script \\ Mathematics and Cryptography }

\author{
(c) The authors, 1998-2002 \\
Frankfurt am Main
}

@@

\parskip 4pt
\vskip + 30 pt
{
In this CrypTool script you will find predominantly mathematically oriented
information for using encryption procedures. The main chapters have been written
by {\em various authors} and are therefore independent from one another. At the end of
most chapters you will find literature and URLs.

You will receive information about the principles of symmetrical and
asymmetrical encryption. A large section of the script is dedicated to the
fascinating topic of prime numbers. Using numerous examples, the elementary number
theory and modular arithmetic are introduced and applied in an exemplary 
manner in the RSA procedure.  After this, one receives insight into the mathematical ideas
behind modern cryptography.

@@

A further chapter is devoted to digital signatures, which are an essential
component of e-business applications. The last chapter --- elliptic curves ---
fits in well with this. The mathematics of elliptic curves forms the basis for
rapid cryptographic algorithms for digital signatures; these algorithms are
extremely well suited for implementation on smartcards.

@@

The {\em authors} Bernhard Esslinger, Bartol Filipovic, Henrik Koy and Roger Oyono
would like to take this opportunity to thank their colleagues in the company and
at the universities of Frankfurt, Giessen and Karlsruhe. They particularly thank
Dr. Peer Wichmann from the Karlsruhe computer science research centre
(Forschungszentrum Informatik, FZI) for his down-to-earth support.

\enlargethispage{0.5cm}
As at CrypTool, the quality of the script is enhanced through your suggestions
and ideas for improvement. We look forward to your feedback.

\vskip +7 pt \noindent
You will find the current version of CrypTool under \newline
  \href{http://www.CrypTool.de}{\tt http://www.CrypTool.de},~~
  \href{http://www.cryptool.com}{\tt http://www.CrypTool.com}~~ or~~
  \href{http://www.cryptool.org}{\tt http://www.CrypTool.org}.
{
\vskip + 7 pt \noindent
The contact persons for this free tool are listed in the readme file for
CrypTool.}
}

>>>>

\title{ CrypTool Script \\ Mathematics and Cryptography \\
Introduction to Elementary Number Theory with Examples}

\author{
(c) B. Esslinger, 1998-2002 \\
Frankfurt am Main \\
besslinger@web.de
}

@@

\parskip 4pt
\vskip + 30 pt
{
This Document contains only the chapter about elementary number theory out of the script
delivered with CrypTool.

Using numerous examples, the elementary number
theory and modular arithmetic are introduced and applied in an exemplary 
manner in the RSA procedure.

@@

@@

The {\em author} 
would like to take this opportunity to thank their colleagues in the company and
at the universities of Frankfurt, Giessen and Karlsruhe. They particularly thank
Dr. Peer Wichmann from the Karlsruhe computer science research centre
(Forschungszentrum Informatik, FZI) for his down-to-earth support.

\enlargethispage{0.5cm}
As at CrypTool, the quality of the script is enhanced through your suggestions
and ideas for improvement. I look forward to your feedback.

\vskip +7 pt \noindent
You will find the current version of CrypTool under \newline
  \href{http://www.CrypTool.de}{\tt http://www.CrypTool.de},~~
  \href{http://www.cryptool.com}{\tt http://www.CrypTool.com}~~ or~~
  \href{http://www.cryptool.org}{\tt http://www.CrypTool.org}.
%{
%\vskip + 7 pt \noindent
%The contact persons for this free tool are listed in the readme file for
%CrypTool.}
}

### 2002-04-15 script-gb/script-gb.tex
no message

In this case the data is encrypted using symmetric procedures: the key is a
session key generated by the sender randomly that is only used for this message.
This session key is then encrypted using the asymmetric procedure and
transmitted to the recipient together with the message. Recipients can determine
the session key using their secret keys and then use the session key to encrypt
the message. In this way, we can benefit from the easy key management \index{Key
management} of asymmetric procedures and encrypt large quantities of data
quickly and efficiently using symmetric procedures.

@@

\subsubsection{GIMPS}\index{GIMP}

@@

\begin{itemize}
  \item The German mathematician \index{Goldbach Christian} Christian Goldbach
(1690-1764) conjectured:
    Every even natural number greater than 2 can be represented as the sum of
two prime numbers.
    Computers have verified\footnote{It is generally accepted today, that
the Goldbach Conjecture is true, i.e. valid for all natural numbers bigger than $2$.
In 1999, mathematician J\"org Richtstein from the computer sciences institute at the University
of Giessen, studied even numbers up to 400 billion and found no contradictory example (see
\href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}).\index{Richstein 1999}
Nevertheless, this does not provide us with general proof.\\
The fact that despite all efforts, Goldbachs' conjecture has to date not been proven leads
one to believe that: since the pioneer work of the Austrian mathematician Kurt G\"odel is well-known,
not every true mathematical theorem is provable (see
\href{http://www.mathematik.ch/mathematiker/goedel.html}{http://www.mathematik.ch/mathematiker/goedel.html}).
Perhaps Goldbachs' Conjecture was correct, despite that however,
proof will never be found. Conversely, that will presumably also remain unproven.}
the Goldbach conjecture for all even numbers up to
$4*10^{14}$ but no general proof has yet been found\footnote{The English
publisher {\em Faber} and the American publisher {\em Bloomsbury} issued in 2000
the 1992 published book ''Uncle Petros and Goldbach's Conjecture'' by Apostolos Doxiadis.
It's the story of an old maths professor who fails to prove a more than 250 old puzzle. To
stimulate the sell rates the English and American publisher compete a prize of 1 million USD, 
is someone proves the conjecture -- to be published till 2004 by an well-known mathematical journal.\\
Susprisingly only British and American citizens are allowed to participate.\\
The theorem closest to Goldbach's conjecture yet was proven 1966 by Chen Jing-Run in a hard to understand fashion:
Each even integer bigger than $2$ is the sum of one prime and of the product of two primes. E.g.: $20=5+3*5.$\\
Most of the research about the Goldbach conjecture is collected in the book: ''Goldbach Conjecture'', ed. Wang Yuan, 1984, World scientific
Series in Pure Maths, Vol. 4.}.
  \item The German \index{Riemann Bernhard} mathematician Bernhard Riemann
(1826-1866) put forward a formula for the distribution of primes that would
further improve the estimate. However, this has neither been proved nor
contradicted so far.
\end{itemize}

@@

All participants who want to \index{X.509}  have X.509 verify that their public
key belongs to a real person consult what is known as a \index{Certification
authority (CA)} certification authority (CA). They prove their identity to this
CA (for example by showing their ID). The CA then issues them an electronic
document (certificate) which essentially contains the names of the certificate-
holder and the CA, the certificate-holder's public key and the validity period
of the certificate. The CA then signs the certificate using its secret key.

@@

Enough about the role of elliptic curves in pure mathematics. In 1985 Neal
Koblitz and Victor Miller independently suggested using elliptic curves in
cryptography. Elliptic curves have thus also found a concrete practical
application. Another interesting field of application for elliptic curves is for
factorising whole numbers. (For example the RAS cryptography system is based on
the \index{Complexity} difficulty/complexity of finding prime factors of an
extremely large number.) In this area, procedures based on elliptic curves have
been investigated and partially used since 1987 (a study by H.W. Lenstra). There
are also prime number tests based on elliptic curves.

>>>>

In this case the data is encrypted using symmetric procedures: the key is a
session key\index{Session key} generated by the sender randomly that is only used for this message.
This session key is then encrypted using the asymmetric procedure and
transmitted to the recipient together with the message. Recipients can determine
the session key using their secret keys and then use the session key to encrypt
the message. In this way, we can benefit from the easy key management \index{Key
management} of asymmetric procedures and encrypt large quantities of data
quickly and efficiently using symmetric procedures.

@@

\subsubsection{GIMPS}\index{GIMPS}

@@

\begin{itemize}
  \item The German mathematician \index{Goldbach Christian} Christian Goldbach
(1690-1764) conjectured:
    Every even natural number greater than 2 can be represented as the sum of
two prime numbers.
    Computers have verified\footnote{It is generally accepted today, that
the Goldbach Conjecture is true, i.e. valid for all even natural numbers bigger than $2$.
In 1999, mathematician J\"org Richtstein from the computer sciences institute at the University
of Giessen, studied even numbers up to 400 billion and found no contradictory example (see
\href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}).\index{Richstein 1999}
Nevertheless, this does not provide us with general proof.\\
The fact that despite all efforts, Goldbachs' conjecture has to date not been proven leads
one to believe that: since the pioneer work of the Austrian mathematician Kurt G\"odel \index{G\""odel Kurt} is well-known,
not every true mathematical theorem is provable (see
\href{http://www.mathematik.ch/mathematiker/goedel.html}{http://www.mathematik.ch/mathematiker/goedel.html}).
Perhaps Goldbachs' Conjecture was correct, despite that however,
proof will never be found. Conversely, that will presumably also remain unproven.}
the Goldbach conjecture for all even numbers up to
$4*10^{14}$ but no general proof has yet been found\footnote{The English
publisher {\em Faber} and the American publisher {\em Bloomsbury} issued in 2000
the 1992 published book ''Uncle Petros and Goldbach's Conjecture'' by Apostolos Doxiadis.
It's the story of an old maths professor who fails to prove a more than 250 old puzzle. To
stimulate the sell rates the English and American publisher compete a prize of 1 million USD, 
is someone proves the conjecture -- to be published till 2004 by an well-known mathematical journal.\\
Susprisingly only British and American citizens are allowed to participate.\\
The theorem closest to Goldbach's conjecture yet was proven 1966 by Chen Jing-Run in a hard to understand fashion:
Each even integer bigger than $2$ is the sum of one prime and of the product of two primes. E.g.: $20=5+3*5.$\\
Most of the research about the Goldbach conjecture is collected in the book: ''Goldbach Conjecture'', ed. Wang Yuan, 1984, World scientific
Series in Pure Maths, Vol. 4.}.
  \item The German \index{Riemann Bernhard} mathematician Bernhard Riemann
(1826-1866) put forward a formula for the distribution of primes that would
further improve the estimate. However, this has neither been proved nor
contradicted so far.
\end{itemize}

@@

All participants who want to \index{X.509}  have X.509 verify that their public
key belongs to a real person consult what is known as a \index{Certification
authority (CA)} certification authority (CA). They prove their identity to this
CA (for example by showing their ID). The CA then issues them an electronic
document (certificate) which essentially contains the names of the certificate-holder
and the CA, the certificate-holder's public key and the validity period
of the certificate. The CA then signs the certificate using its secret key.

@@

Enough about the role of elliptic curves in pure mathematics. In 1985 Neal
Koblitz and Victor Miller independently suggested using elliptic curves in
cryptography. Elliptic curves have thus also found a concrete practical
application. Another interesting field of application for elliptic curves is for
factorising whole numbers. (For example the RSA cryptography system is based on
the \index{Complexity} difficulty/complexity of finding prime factors of an
extremely large number.) In this area, procedures based on elliptic curves have
been investigated and partially used since 1987 (a study by H.W. Lenstra). There
are also prime number tests based on elliptic curves.
